---
title: "How to use malt?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use malt?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(malt)
```

Goal: generate samples $X_1,\cdots,X_n\in\mathbb{R}^d$ approximately distributed from
$$
\Pi(x)\propto e^{-\Phi(x)}
$$
by evaluating a potential function $\Phi$ and its gradient.

Example 1: Gaussian
$$
\begin{aligned}
\Phi(x)&=\frac{1}{2}(x-\mu)^\top\Sigma^{-1}(x-\mu)\\
\nabla\Phi(x)&=\Sigma^{-1}(x-\mu)
\end{aligned}
$$
Suppose we want to sample from a Gaussian distribution with heterogeneous scales, such that
$$
\mu=0_d,\qquad \Sigma=\underset{1\le i\le d}{\rm diag}(\sigma_i^2),\qquad \sigma_i^2=i/d,\qquad d=50.
$$
We specify the corresponding potential function and its gradient
```{r}
d=100
sigma=((d:1)/d)^(1/2)
potential=function(x){sum(0.5*x^2/sigma^2)}
grad=function(x){x/sigma^2}
```

We choose a friction, time step and integration time. We run malt from a deterministic initialization and draw $n=10000$ samples. We also run hmc (friction: g=0) to compare. Both outputs are stored.
```{r}
g=1.5
h=0.15
T_phys=1.6
x_init=rnorm(d)*sigma
x_init=rep(5,d)
n=10^4
output_malt=malt(potential,grad,n,g,h,T_phys,x_init)
output_hmc=malt(potential,grad,n,0,h,T_phys,x_init)
```
We obtain the acceptance rates of malt and hmc.
```{r}
output_malt$acceptance_rate
output_hmc$acceptance_rate
```

We represent graphically the traceplots and ACFs for the first component
```{r, out.width='.50\\linewidth', fig.width=3, fig.height=3,fig.show='hold'}
par(mar=c(4,4,1,1))
index=1
plot(output_malt$chain[1:500,index],type="l",xlab="iterations",ylab=paste("traceplot n°",index," - MALT"))
plot(output_hmc$chain[1:500,index],type="l",xlab="iterations",ylab=paste("traceplot n°",index," - HMC"))
acf(output_malt$chain[,index],ylab=paste("ACF n°",index," - MALT"),main="")
acf(output_hmc$chain[,index],ylab=paste("ACF n°",index," - HMC"),main="")
summary_malt=ess(output_malt)
summary_hmc=ess(output_hmc)
table_ess=matrix(c(
summary_malt$ess_mean[index],
summary_hmc$ess_mean[index],
summary_malt$ess_square[index],
summary_hmc$ess_square[index]),nr=2)
rownames(table_ess)=c("MALT","HMC")
colnames(table_ess)=c("ESS mean","ESS square")
table_ess
```

We illustrate traceplots and ACFs for another component (quantile 2% worst positive autocorrelation for hmc)
```{r, out.width='.50\\linewidth', fig.width=3, fig.height=3,fig.show='hold'}
library(knitr)
par(mar=c(4,4,1,1))
index=which(rank(summary_hmc$ess_mean)==floor(d/50))[1]
plot(output_malt$chain[1:500,index],type="l",xlab="iterations",ylab=paste("traceplot n°",index," - MALT"))
plot(output_hmc$chain[1:500,index],type="l",xlab="iterations",ylab=paste("traceplot n°",index," - HMC"))
acf(output_malt$chain[,index],ylab=paste("ACF n°",index," - MALT"),main="")
acf(output_hmc$chain[,index],ylab=paste("ACF n°",index," - HMC"),main="")
table_ess=matrix(c(
  summary_malt$ess_mean[index],
  summary_hmc$ess_mean[index],
  summary_malt$ess_square[index],
  summary_hmc$ess_square[index]),nr=2)
rownames(table_ess)=c("MALT","HMC")
colnames(table_ess)=c("ESS mean","ESS square")
table_ess
#kable(table_ess,row.names=c("MALT","HMC"),col.names=c("ESS mean","ESS square"))
```

We illustrate traceplots and ACFs for another component (quantile 2% worst negative autocorrelation for hmc)
```{r, out.width='.50\\linewidth', fig.width=3, fig.height=3,fig.show='hold'}
par(mar=c(4,4,1,1))
index=which(rank(summary_hmc$ess_square)==floor(d/50))[1]
plot(output_malt$chain[1:500,index],type="l",xlab="iterations",ylab=paste("traceplot n°",index," - MALT"))
plot(output_hmc$chain[1:500,index],type="l",xlab="iterations",ylab=paste("traceplot n°",index," - HMC"))
acf(output_malt$chain[,index],ylab=paste("ACF n°",index," - MALT"),main="")
acf(output_hmc$chain[,index],ylab=paste("ACF n°",index," - HMC"),main="")
table_ess=matrix(c(
  summary_malt$ess_mean[index],
  summary_hmc$ess_mean[index],
  summary_malt$ess_square[index],
  summary_hmc$ess_square[index]),nr=2)
rownames(table_ess)=c("MALT","HMC")
colnames(table_ess)=c("ESS mean","ESS square")
table_ess
```


We illustrate the distribution of the ESS when estimating the mean and the square.
```{r, out.width='.50\\linewidth', fig.width=3, fig.height=3,fig.show='hold'}
boxplot(summary_malt$ess_mean,summary_hmc$ess_mean,names=c("MALT","HMC"),ylim=c(0,n),main="ESS mean")
boxplot(summary_malt$ess_square,summary_hmc$ess_square,names=c("MALT","HMC"),ylim=c(0,n),main="ESS square")
summary(summary_malt$ess_mean)
summary(summary_hmc$ess_mean)
summary(summary_malt$ess_square)
summary(summary_hmc$ess_square)
```



